Concurrent Programming with CXL

### 2020-05-21 book/paper.tex
Latest

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  In order to resolve this, we
will model processes in critical regions using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical region more than once, but it can also choose to terminate, even
without entering the critical region ever.
Note also that we have dropped the \texttt{@ncs} label: it would serve no
specific purpose.

\begin{quote}
Try it out!
\begin{itemize}
\item Plug \texttt{while True: pass;;} for entering the critical region
in Figure~\ref{fig:cs} and run CXL.  It should print a trace
to a state from which a terminating state cannot be reached.
\end{itemize}
\end{quote}

@@

\chapter{Peterson's Algorithm}

@@

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
The program shows two kinds of processes: \texttt{reader()} and
\texttt{writer()}.  The critical section is spread between two
labels: readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.

A reader process that wants to enter \texttt{@rcs} first waits
until there are no writers: \texttt{nwriters == 0}.
If so, it increments the number of readers.  Similarly, a writer
process waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = n$,
\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} = n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.

@@

Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
that does not busy-wait.
It uses two locks: \texttt{rwlock} is used by both readers and writers,
while \texttt{rlock} is only used by readers.

>>>>

In order to detect violations of progress, and other liveness problems in
algorithms in general, CXL requires that every execution must be
able to reach a state in which all processes have terminated.
Clearly, even if mutual exclusion holds in Figure~\ref{fig:csbarebones},
the spawned processes never terminate.  In order to resolve this, we
will model processes in critical sections using the framework in
Figure~\ref{fig:cs}: a process can \emph{choose} to enter a
critical section more than once, but it can also choose to terminate, even
without entering the critical section ever.
Note also that we have dropped the \texttt{@ncs} label: it would serve no
specific purpose.

\begin{quote}
Try it out!
\begin{itemize}
\item Plug \texttt{while True: pass;;} for entering the critical section
in Figure~\ref{fig:cs} and run CXL.  It should print a trace
to a state from which a terminating state cannot be reached.
\end{itemize}
\end{quote}

@@

\chapter{Peterson's Algorithm}
\label{ch:peterson}

@@

Figure~\ref{fig:rwbusy} presents a solution that uses a
single (ordinary) lock and two counters: one that maintains the number
of readers and one that maintains the number of writers.
The lock is used to protect access to those counters.
The program shows two kinds of processes: \texttt{reader()} and
\texttt{writer()}.  The critical section is spread between two
labels: readers access \texttt{@rcs} and writers access \texttt{@wcs}.
The specification is that if a reader is at label \texttt{@rcs},
no writer is allowed to be at label \texttt{@wcs}.  Vice versa, if
a writer is at label \texttt{@wcs}, no reader is allowed to be at
label \texttt{@rcs} \emph{and} there cannot be another writer at
label \texttt{@wcs}.

A reader process that wants to enter \texttt{@rcs} first waits
until there are no writers: \texttt{nwriters == 0}.
If so, it increments the number of readers.  Similarly, a writer
process waits until there are no readers \emph{or} writers.
If so, it increments the number of writers.
The important invariants in this code are:
\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $n$ writers at \texttt{@rcs} $\Rightarrow \mathtt{nwriters} \ge n$,
\item $(\mathtt{nreaders} \ge 0 \land \mathtt{nwriters} = 0) \lor
    (\mathtt{nreaders} = 0 \land 0 \le \mathtt{nwriters} \le 1)$.
\end{itemize}
It is easy to see that the invariants hold and imply the reader/writer
specification.
The solution also supports progress: if no process is in the critical
section then any process can enter.  Better still: if any reader is in the
critical section, any other reader is also able to enter.

@@

Figure~\ref{fig:rw2lock} presents a reader/writer lock implementation
that does not busy-wait.
It uses two locks: \texttt{rwlock} is used by both readers and writers,
while \texttt{rlock} is only used by readers.
\texttt{rwlock} is held either when readers are at label \texttt{@rcs}
or when a writer is at label \texttt{@wcs}.
\texttt{rlock} is used to protect the \texttt{nreaders} variable that
counts the number of readers in the critical section.

The invariants that imply the reader/writer specification
(which are again easy to verify) are as follows:

\begin{itemize}
\item $n$ readers at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} \ge n$,
\item $\exists$ writer at \texttt{@rcs} $\Rightarrow \mathtt{nreaders} = 0$,
\end{itemize}

A writer simply acquires \texttt{rwlock} to enter the critical section
and releases it to exit.  The \emph{first} reader to enter the critical
section acquires \texttt{rwlock} and the \emph{last} reader to exit
the critical section releases \texttt{rwlock}.
The implementation satisfies progress: if no process is in the critical
section than any process enter.

It is instructive to see what happens when a writer is in the critical
section and two readers try to enter.  The first reader successfully
acquires \texttt{rlock} and but \emph{hangs} when trying to acquire
\texttt{rwlock}, which is held by the writer.  The second reader hangs
trying acquire \texttt{rlock} because it is held by the first reader.
When the writer leaves the critical section, the first reader acquires
\texttt{rwlock}, sets \texttt{nreaders} to 1, and releases \texttt{rlock}.
\texttt{rwlock} is still held.
Then the second reader acquires \texttt{rlock} and, assuming the first
reader is still in the critical section, increments \texttt{nreaders} to 2
and enter the critical section \emph{without} acquiring \texttt{rwlock}.
\texttt{rwlock} is essentially jointly held by both readers.
It does not matter in which order they leave: the second will release
\texttt{rwlock}.

\chapter{Starvation}

So far we have pursued two properties: emph{mutual exclusion}
and \emph{progress}.  The former is an example of
\texttt{safety property}---it prevents something ``bad'' from
happening, like a reader and writer process both entering the
critical section.  The \emph{progress} property is an example
of a liveness property---that something good eventually happens.
Informally (and inexactly), progress states that if no processes
are in the critical section, then some process that wants to enter
can.

Progress is a weak form of liveness.  It says that \emph{some}
process can enter, but it does not prevent a scenario such as
the following.  There are three processes repeatedly trying to
enter a critical section using a spinlock.  Two of
the processes successfully enter, alternating, but the third
process never gets a turn.  This is an example of
\texttt{starvation}.  With a spinlock, this scenario could
even happen with two processes.  Initially both processes
try to acquire the spinlock.  One of the processes is
successful and enter.  After the process leaves, it immediately
tries to re-enter.  This state is identical to the initial
state, and there is nothing that prevents the same process
from acquiring the lock yet again.

It is worth noting that Peterson's Algorithm (Chapter~\ref{ch:peterson})
does not suffer from starvation, thanks to the \texttt{turn} variable
that alternates between 0 and 1 when two processes are contending for
the critical section.

While spinlocks suffer from starvation, it is a uniform random
process and each process has an equal chance of entering the critical
section.  Thus the probability of starvation is exponentially vanishing.
Unfortunately, such is not the case for the
reader/writer solution of Figure~\ref{fig:rw2lock}.
Consider this scenario: there are two readers and one writer.  One reader
is in the critical section while the writer is waiting.  Now the
second reader tries to enter and is able to.  The first reader leaves.
We are now in a similar situation as the initial state with one reader
in the critical section and the writer waiting, but it is not the same
reader.  Unfortunately for the writer, this scenario can repeat itself
indefinitely.  So even if neiter reader was in the critical section
for all of the time, the writer never had a chance.
In this section we will present a version of a reader/writer lock
implementation that solves this type of starvation.
